---
title: "1_Descriptive_Statistics"
author: Mar Batlle
output: md_document
---

## Load libraries
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Load libraries
library(knitr)
library(tinytex)
library(tidyverse)
library(ggplot2)
library(DataExplorer)
library(nortest)
```

## The dataset: Low weight at birth
**Goal of the study**: establish which factors are associated with low weight on newborns, present in women who have given birth.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Read the data
low_weight_births <- read.csv("~/Documentos/bioestadistica/datos/Bajo peso al nacer.csv", sep=";") %>%
  data.frame()
# Let's get an idea of what we're working with 
glimpse(low_weight_births)
```

As you can see, the set of variables present in this dataset contains are described in english, to make it easier to understand, here you can see the meaning of each variable and its corresponding categories:
* **id**: identification
* **bajo_pes**: low weight at birth (categories: 0 = normal weight (>= 2500g), 1 = low weight (< 2500g))
* **edad**: age
* **peso**: weight
* **raza**: race (categories: 1 = white, 2 = black, 3 = other)
* **fumador**: smoker (categories: 0 = no, 1 = yes)
* **part_pre**: premature birth (categories: 0 = no, 1 = 1 birth, 2 = 2 births, 3 = more than 3)
* **hta**: hipertension (categories: 0 = no, 1 = yes)
* **irr_urin**: urinary irritability (categories: 0 = no, 1 = yes)
* **visi_med**: number of medical visits

To condinue with the analysis, we we'll clean the data while modifying it's variable names and categories.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Clean and adequate the data
data  <- low_weight_births %>%
  drop_na() %>%
  rename(low_weight = bajo_pes,
         age = edad,
         weight = peso,
         race = raza,
         smoker = fumador,
         prem_birth = part_pre,
         urin_irr = irr_urin,
         med_vis = visi_med) %>%
  mutate(low_weight = factor (low_weight, labels = c("normal","low")),
         race = factor (race, labels = c("white","black","others")),
         smoker = factor (smoker, labels = c("no", "yes")),
         prem_birth = factor (prem_birth, labels=c("no","1_birth","2_births","more2_births")),
         hta = factor (hta, labels=c("no","yes")),
         urin_irr = factor (urin_irr, labels=c("no","yes")))

glimpse(data)
```


# Descriptive Statistics


## Single Variable Analysis

### Description of a Discrete Variable
* Number of predetermined finite categories. Alphabetic or numeric codes
* Describe a quality of the sample
* Ordinal variables: the categories have a natural order

#### Summary Measures
As an example and guide, we analize the response variable; Bajo peso al nacer:

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
low_weight <- data$low_weight

# Absolute frequencies: Number of observations presented by each of the categories
t1 <- table(low_weight)
t1

# # Relative frequencies: Quotient between absolute frequencies and sample size, expressed in proportions or percentages
p1 <- prop.table(t1)
p1_round <- round(p1*100, dig=2)
p1_round
```

Using the DataExplorer package, we can easily plot the frequency distribution of all discrete variables:
```{r}
plot_bar(data)
```

#### Graphics
As in the previous example, we analize the response variable; *low weight at birth*, as a guide. Also, have in mind that to execute graphs in a terminal, we always need to write the function dev.new() first
```{r}
# Sectors diagrams: Graphic representation in the form of a circle with areas proportional to theabsolute or relative frequency of each category
dev.new()
par(mfrow=c(1,2))
pie(t1)
pie(t1, col=c("red","blue"), main="Low weight at birth", 
        labels=paste(names(t1), " (", p1_round ,"%)", sep=""))
```
```{r}
# Bar chart: Graphical representation where the abscissa axis represents the categories and relative or absolute frequencies on the ordinate axis
dev.new()
par(mfrow=c(1,2))
barplot(t1)
barplot(t1, col=c("red","blue"), main="Low weight at birth", 
        names.arg=paste(names(t1), " (", round( p1 * 100, dig=2),"%)", sep=""))
```


### Description of a Continuous Variable
* Variables measured on a numerical scale
* Defined range: minimum and maximum

#### Summary measures of central tendency
As an example and guide, we analize the variable *age*^.
* *Mean:* This measures describes the arithmetic  of the sample:
```{r}
mean(data$age)
```
* *Median:* Central value in an ordered data set, half data on the right and half on the left. 
```{r}
median(data$age)
```
The median is a robust estimator because it is not as sensitive as the mean to distant values.

#### Position or location measurements
The quantile or percentile of order p of a distribution (0 <p <1): Pp value such that a proportion p of data values <= Pp.

* Quartiles: percentiles 0.25, 0.50, 0.75
* Tertiles: 0.3333, 0.6666 percentiles
* Deciles: percentiles 0.10, 0.20,…., 0.90

```{r}
quantile(data$age, c(0.10,0.25,0.50,0.75,0.90))
```

The median and the quartiles Q1 and Q3 divide the observations into 4 parts that they contain 25% of the data each. The mean is not enough to describe a sample, a measure of dispersion is necessary.

#### Measures of dispersion
Measures that quantify how scattered the data is around of a measure of central tendency. Its important to mention that the following functions contain an important parameter to treat missing values; na.rm=TRUE.

* *Variance:* It is the mean of the squared deviations from the mean
```{r}
var(data$age)
```
* *Typical deviation:* It is the square root of the variance, and is a measure of variability in the same units as the sample data.
```{r}
sd(data$age)
```
* *Coefficient of variation:* Express variability in a unitless measure.
```{r}
100 * abs(sd(data$age) / mean(data$age))
```
* *Rangs*: is the difference between the maximum and minimum value.
```{r}
range(data$age)
```
* *Interquartile range*: is the difference between the first and third quartiles, and is a measure of dispersion in 50% of the core data.
```{r}
IQR(data$age)
```
* *MAD (median absolute dev*iation): is the median of the deviations in value absolute of the observations with respect to the median.
```{r}
mad(data$age)
```
* The interquartile range and MAD are robust estimators because they are not as sensitive as the standard deviation to far values, since they are based on in the median and in the position measures.

#### Graphics

* *Histogram:* Bar chart where the frequencies of the grouped variable are represented in intervals
```{r}
par(mfrow=c(1,2))
hist(data$age)
hist(data$age, col="red", main="Histogram of Age", 
              xlab="Age", ylab="Frequency",
              xlim=c(10,50), breaks=seq(10,50, by=2))
```
* *Boxplot:* Graphic based on the distribution of the quartils (including median), with an easy identification of the outliers values.
```{r}
par(mfrow=c(1,2))
boxplot(data$age)
boxplot(data$age, col="red", main="Boxplot of Age", xlab="Age")
```

Using the DataExplorer package, we can easily and quickly create histograms and density plots to analyze all continuous variables present in the study.
```{r}
# View histogram of all continuous variables
plot_histogram(data)

# View estimated density distribution of all continuous variables
plot_density(data)
```

### Summary Function
The summary () function has as an argument a data frame, and displays:
* Absolute frequencies for categorical variables
* Basic statistics for continuous variables
* Number of missings of each variable, if any
```{r}
summary(data)
```

## Normality
Some statistical tests require that the variables be distributed according to a normal distribution, and this must be checked

### Checking normality. Hypothesis test
* H0: the data are distributed according to a normal distribution
* Shapiro-Wilks test: Small samples (<30, <50)
```{r}
# Our sample is larger than 50, so this wouldn't be the best choice, but we'll still see the command.
shapiro.test(data$age)
```
* Kolmogorov-Smirnov test with correction: Large samples (> 30,> 50). It is a more general test, and it can be used for any distribution besides the normal - needs to load nortest library
```{r}
lillie.test(data$age)
```
* In both cases we reject the hypothesis of normality for p-values < 0.05

### Checking normality. Graphical methods
* Q-Q Plot: the quantiles observed in the data are shown against the theoretical ones; If the points fit a line, the data follow the theoretical distribution

```{r}
qqnorm(data$age, main="Normal Q-Q Plot - Age")
qqline(data$age)
```

Using the DataExplorer package, we can easily and quickly create qq plots to analyse the normality of all continuous variables present in the study.
```{r}
# View quantile-quantile plot of all continuous variables
plot_qq(data)
```

### Transformations of a variable
* A continuous variable can be transformed to make it follow a normal distribution, or at least have a symmetric distribution. The family of Transformations Power Xp is usually used with p ≠ 0. It is useful to try a simpler and more interpretable set of these transformations
* If we want to correct the skew to the right, we increase the power (x²); sqrt(max(x+1)-x),log10(max(x+1)-x)
* If we want to correct the skew to the left, we dicrease the power; sqrt(x), log10(x)

Interesting resource for data transformations: https://www.datanovia.com/en/lessons/transform-data-to-normal-distribution-in-r/


```{r echo=FALSE}
# Session Info
sessionInfo()
```
